---
title: "Consumer Behavior Analysis"
author: "Elizabeth Kinuthia"
date: "2023-05-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Consumer Behavior Analysis

In this project, I analyzed Cyclistic's company data in order to answer stakeholder questions. I performed EDA, statistical analysis, and created visualizations to drive decision making.

## Setting Up My Environment

I began setting up my enviroment by loading the necessary packages that I will use during my analysis.

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(janitor)
```

## Preparing My Data and Environment for Analysis

I then collected my data, and loaded all the data into a dataframe. After that, I cleared up space in my environment to ensure my environment is as efficient as possible. 

```{r}
#collect the Data
trips_05_2022 <- read_csv('202205-divvy-tripdata.csv')  
trips_06_2022 <- read_csv('202206-divvy-tripdata.csv')
trips_07_2022 <- read_csv('202207-divvy-tripdata.csv')
trips_08_2022 <- read_csv('202208-divvy-tripdata.csv')
trips_09_2022 <- read_csv('202209-divvy-publictripdata.csv')
trips_10_2022 <- read_csv('202210-divvy-tripdata.csv')
trips_11_2022 <- read_csv('202211-divvy-tripdata.csv')
trips_12_2022 <- read_csv('202212-divvy-tripdata.csv')
trips_01_2023 <- read_csv('202301-divvy-tripdata.csv')
trips_02_2023 <- read_csv('202302-divvy-tripdata.csv')
trips_03_2023 <- read_csv('202303-divvy-tripdata.csv')
trips_04_2023 <- read_csv('202304-divvy-tripdata.csv')

#combine all the data into one dataframe
alltrips <- bind_rows(trips_05_2022, trips_06_2022,trips_07_2022, 
                      trips_08_2022, trips_09_2022, trips_10_2022, 
                      trips_11_2022, trips_12_2022, trips_01_2023, 
                      trips_02_2023, trips_03_2023, trips_04_2023)

#clear up space in the environment
remove(trips_05_2022, trips_06_2022,trips_07_2022, 
       trips_08_2022, trips_09_2022, trips_10_2022, 
       trips_11_2022, trips_12_2022, trips_01_2023, 
       trips_02_2023, trips_03_2023, trips_04_2023)

#create new dataframe to add new columns
cyclistic_df <- alltrips 
```

## Data Wrangling

In this section, I performed data wrangling by transforming my data, cleaning it, and validating the data using the `skim_without_charts()` function. 

### Transforming the Data

The first thing that we need to do is determine the ride length. I created a column for `ride_length` 

```{r}
cyclistic_df$ride_length <- difftime(cyclistic_df$ended_at, cyclistic_df$started_at)
```

Another problem with our data is that it is too granular. The data can only be aggregated at a ride level, so it's necessary to add more columns that provide additional opportunities to aggregate the data. I created columns for date, day of week, month, year, time, and hour. 

```{r}
cyclistic_df$date <- as.Date(cyclistic_df$started_at)
cyclistic_df$day_of_week <- format(as.Date(cyclistic_df$started_at), '%A')
cyclistic_df$day <- format(as.Date(cyclistic_df$started_at), '%d')
cyclistic_df$month <- format(as.Date(cyclistic_df$started_at), '%m')
cyclistic_df$year <- format(as.Date(cyclistic_df$started_at), '%Y')
cyclistic_df$time <- format(as.Date(cyclistic_df$started_at), '%H:%M:%S')
cyclistic_df$hour <- hour(cyclistic_df$started_at)
```

I also created a column to indicate the season, i.e. Spring, Summer, Autumn, and Winter. 

```{r}
cyclistic_df <- cyclistic_df %>% mutate(season=
                                          case_when(month == '03'~'Spring',
                                                    month == '04'~'Spring',
                                                    month == '05'~'Spring',
                                                    month == '06'~'Summer',
                                                    month == '07'~'Summer',
                                                    month == '08'~'Summer',
                                                    month == '09'~'Autumn',
                                                    month == '10'~'Autumn',
                                                    month == '11'~'Autumn',
                                                    month == '12'~'Winter',
                                                    month == '01'~'Winter',
                                                    month == '02'~'Winter'))
```

I also created a column to indicate the time of day. 

```{r}
cyclistic_df <- cyclistic_df %>% mutate(time_of_day=
                                          case_when(hour == '0'~'Night',
                                                    hour == '1'~'Night',
                                                    hour == '2'~'Night',
                                                    hour == '3'~'Night',
                                                    hour == '4'~'Night',
                                                    hour == '5'~'Night',
                                                    hour == '6'~'Morning',
                                                    hour == '7'~'Morning',
                                                    hour == '8'~'Morning',
                                                    hour == '9'~'Morning',
                                                    hour == '10'~'Morning',
                                                    hour == '11'~'Morning',
                                                    hour == '12'~'Afternoon',
                                                    hour == '13'~'Afternoon',
                                                    hour == '14'~'Afternoon',
                                                    hour == '15'~'Afternoon',
                                                    hour == '16'~'Afternoon',
                                                    hour == '17'~'Evening',
                                                    hour == '18'~'Evening',
                                                    hour == '19'~'Evening',
                                                    hour == '20'~'Evening',
                                                    hour == '21'~'Evening',
                                                    hour == '22'~'Night',
                                                    hour == '23'~'Night'))
```

### Data Cleaning

I first used `skim_without_charts()` to check which columns had null values in them. 

```{r}
skim_without_charts(cyclistic_df)
```

I then I cleaned the data by removing unnecessary columns, null values and duplicate rows. 

```{r}
cyclistic_df <- cyclistic_df %>% 
  select(-c(ride_id, start_station_id, end_station_id, start_lat, start_lng, end_lat, end_lng)) 

cyclistic_df <- cyclistic_df %>% 
  filter(!(is.na(started_at))) 

cyclistic_df <- distinct(cyclistic_df)
```

Then, I filtered out all trips with a negative value for trip duration, and the trips from start station 'HQ QR'. This is because these bikes were taken out of circulation by Cyclistic for quality control reasons.

```{r}
cyclistic_df <- cyclistic_df %>% 
  filter(!(start_station_name == 'HQ QR' | ride_length <0))
```

I then set the order for these columns: 

  1) `day_of_week` column
  2) `time_of_day` column
  3) `season` column

```{r}
#set order for day_of_week column
cyclistic_df$day_of_week <- cyclistic_df$day_of_week %>% 
  ordered(levels = c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday'))

#set order for time_of_day column
cyclistic_df$time_of_day <- cyclistic_df$time_of_day %>% 
  ordered(levels = c('Morning','Afternoon','Evening','Night'))

#set order for season column
cyclistic_df$season <- cyclistic_df$season %>% 
  ordered(levels = c('Winter','Spring','Summer','Autumn'))
```

Finally, I viewed my data to check the quality of the data after performing data wrangling. 

```{r}
skim_without_charts(cyclistic_df)
```

## Exploratory and Statistical Data Analysis

In this phase, I performed Exploratory Data Analysis and Statistical Analysis, and used visualizations to analyze my data. 

### Analyzing Total Rides
I found the total number of rides and visualized it according to user type using the `ggplot2` library. 
```{r}
nrow(cyclistic_df)
```

```{r}
cyclistic_df %>% 
  group_by(member_casual) %>% 
  count(member_casual)
```

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=member_casual))
```

I analyzed the total rides per member type and per bike type.

```{r}
cyclistic_df %>% 
  group_by(member_casual) %>% 
  count(rideable_type)

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=member_casual, fill=rideable_type))
  
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=member_casual)) + facet_wrap(~rideable_type)
```

I analyzed the total rides per day of the week

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=day_of_week)) + facet_wrap(~member_casual)

cyclistic_df %>% 
  group_by(day_of_week, member_casual) %>% 
  count()
```

I analyzed the total rides per month

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=month)) + facet_wrap(~member_casual)

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=month))

cyclistic_df %>% 
  group_by(month) %>% 
  count()
```

I analyzed the total rides per year

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=year)) + facet_wrap(~member_casual)

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=year))

cyclistic_df %>% 
  group_by(year) %>% 
  count(member_casual)
```

I analyzed the total rides per season

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=season, fill = member_casual))

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=season))

cyclistic_df %>% 
  group_by(season) %>% 
  count(member_casual)
```

I analyzed the total rides for each time of day

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=time_of_day, fill = member_casual))

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=time_of_day))

cyclistic_df %>% 
  group_by(time_of_day) %>% 
  count(member_casual)
```

I analyzed the total rides per hour

```{r}
cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=hour, fill = member_casual))

cyclistic_df %>% 
  ggplot() + geom_bar(aes(x=hour))

cyclistic_df %>% 
  group_by(hour) %>% 
  count() %>% 
  max()
```

### Analyzing Trip Length

First, I created a column to be able to analyze the trip length in minutes. 

```{r}
cyclistic_df$ride_length_min <- difftime(cyclistic_df$ended_at, cyclistic_df$started_at, units = "mins")
```

Then I calculated the mean, and viewed the distribution of the trip length. 

```{r}
mean(as.numeric(cyclistic_df$ride_length_min)) #average trip length in minutes
summary(as.numeric(cyclistic_df$ride_length_min)) #distribution of trip length
```

I analyzed the average trip length per user type.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=member_casual, y=as.numeric(ride_length_min))) + geom_bar(stat = "summary",fun="mean")

cyclistic_df %>% 
  group_by(member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length per bike type.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=rideable_type, y=as.numeric(ride_length_min))) + 
  geom_bar(stat = "summary",fun="mean") +
  facet_wrap(~member_casual)

cyclistic_df %>% 
  group_by(rideable_type, member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length per month.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=month, y=as.numeric(ride_length_min))) + geom_bar(stat = "summary",fun="mean")

cyclistic_df %>% 
  group_by(month) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length per year. 

```{r}
cyclistic_df %>% 
  ggplot(aes(x=year, y=as.numeric(ride_length_min))) + geom_bar(stat = "summary",fun="mean")

cyclistic_df %>% 
  group_by(year) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length per season.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=season, y=as.numeric(ride_length_min))) + geom_bar(stat = "summary",fun="mean")

cyclistic_df %>% 
  group_by(season) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length for each time of day.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=time_of_day, y=as.numeric(ride_length_min),fill=member_casual)) + 
  geom_bar(stat = "summary",fun="mean")

cyclistic_df %>% 
  group_by(time_of_day,member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()
```

I analyzed the average trip length per hour.

```{r}
cyclistic_df %>% 
  ggplot(aes(x=hour, y=as.numeric(ride_length_min))) + 
  geom_bar(stat = "summary",fun="mean") +
  facet_wrap(~member_casual)

cyclistic_df %>% 
  group_by(hour,member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length_min)) %>% 
  as.data.frame()

cyclistic_df %>% 
  group_by(hour) %>% 
  summarise(mean_ride_length = mean(as.numeric(ride_length_min))) %>% 
  as.data.frame() %>% max()
```

After I had performed my analysis I was able to gain the following insights:

  1) Even though annual members took the most bike rides, casual riders rode their bikes for longer periods of time each ride. 
    
    - Casual riders took their longest bike rides at night. 

  2) The classic bike was the most popular bike among annual members, while the electric bike was most popular among casual riders.
    
    - Casual riders rode the longest on the docked bikes, even though docked bikes had the least amount of rides total. 

  3) Casual riders prefer to ride their bikes on weekends, where as annual members ride their bikes most during weekdays. 

  4) Summer was the most popular season for bike riding for both annual and casual riders.

## Data Visualization Preparation
 
Now that the analysis was complete, I prepared the data for visualization in Tableau. 

```{r}
cyclistic_viz <- cyclistic_df
cyclistic_viz <- cyclistic_viz %>% mutate(month = 
                                                case_when(month == '01'~'January',
                                                          month == '02'~'February',
                                                          month == '03'~'March',
                                                          month == '04'~'April',
                                                          month == '05'~'May',
                                                          month == '06'~'June',
                                                          month == '07'~'July',
                                                          month == '08'~'August',
                                                          month == '09'~'September',
                                                          month == '10'~'October',
                                                          month == '11'~'November',
                                                          month == '12'~'December')) 
```

I made sure to set the order for the month column. 

```{r}
cyclistic_viz$month <- cyclistic_viz$month %>% 
  ordered(levels = c('January','February','March','April','May','June','July',
                     'August','September','October','November','December'))
```

I then exported my data to a csv file. With that, my analysis using R programming was complete. 

```{r}
library(data.table)
fwrite(cyclistic_viz,'cyclistic_tableau_data.csv')
```